### 2020-09-28 latex/acrobot.tex
Finished writing the first version of the acrobot dynamics

\begin{figure}
    \centering
    \includestandalone[width=0.7\textwidth]{images/acrobot_model}
    \caption{The acrobot model.}%
    \label{fig:acrobot-model}
\end{figure}

@@

\section{Dynamics of the Acrobot}
% TODO: Derive the acrobot dynamics in Hamiltonian form
For the acrobot model displayed in Figure \ref{fig:acrobot-model}, 
the COM of each link lies somewhere along the link itself. 
While this is the most general representation of an acrobot, the dynamics are
somewhat complicated and are difficult to use when proving the results in the 
rest of this chapter. For simplicity, we will
assume the acrobot is comprised of two massless rods of equal length with 
equal masses at the tips, as in Figure \ref{fig:simple-acrobot-model}; we call
this a \textbf{simple} acrobot.

\begin{figure}
    \centering
    \includestandalone[width=0.5\textwidth]{images/simple_acrobot_model}
    \caption{A simple acrobot with lengths \(l\) and masses of size \(m\).}
    \label{fig:simple-acrobot-model}
\end{figure}

\section{Previous Approaches}
% TODO: Give Xingbo's intuition, abstract away to qa = sin(theta) constraint
% based on VLP but show it doesn't work well in our framework
% TODO: Describe Xingbo's results and observe that his results were similar to a VNHC
% except they used VHC tools. We will modify his approach to use VNHCs and prove
% rigorously that we can stabilize any energy level set on the acrobot.

@@

%\section{Energy Injection for the Acrobot}
%Suppose the acrobot is simple; 
%that is, assume that the acrobot's two links are of equal length \(l\)
%and of equal mass \(m\), with the mass concentrated at the tip of each rod.
%
%Let \(q = (q_u,q_a)\) where \(q_u\) is the shoulder angle and \(q_a\) is the hip angle.
%
%Then the acrobot has hamiltonian given by
%\[
%    \mathcal{H}(q,p) = \frac{1}{2} p^T M^{-1}(q) p + V(q)
%\]
%where
%\[
%    M^{-1}(q) = \frac{1}{m l^2 (2 - \cos^2(q_a))} 
%    \begin{bmatrix}
%        1              & -(1 + \cos(q_a)) \\
%        -(1 + \cos(q_a)) & (3 + 2\cos(q_a))
%    \end{bmatrix}
%\]
%and
%\[
%    V(q) = m g l (3 - 2 \cos(q_u) - \cos(q_u + q_a))
%\]
%
%Notice that the acrobot satisfies \(\partial M^{-1}(q)/ \partial q_u = 0\),
%which means it may be possible to enforce a regular solvable VNHC to inject energy
%into the acrobot.
%One such VNHC is
%\begin{align}\label{eqn:acrobot_vnhc}
%    \begin{split}
%    h(q,p) &= q_a - f(q_u,p_u) \\
%           &= q_a - \bar{q_a}\sin(\arctan(p_u,q_u)) \\
%           &= q_a - \frac{\bar{q_a}p_u}{\sqrt{q_u^2 + p_u^2}}
%    \end{split}
%\end{align}
%where \(\bar{q_a} \in ]0,\pi[\) is a parameter which dictates how much the 
%controlled link can swing back and forth.
%
%To compute the regularity and solvability conditions, first observe that since
%the acrobot has two coordinates and only \(p_a\) is actuatable, the control matrix
%\(B = [0, 1]^T\) is constant and full rank everywhere. This also implies that
%\(h\) is both regular and solvable if \(dh_q M^{-1}(q)B\) is full rank.
%
%Observe that
%\begin{align}
%    \begin{split}
%        dh_q M^{-1}(q)B &= \frac{1}{m l^2 (2 - \cos^2(q_a))}
%\left[ \frac{\bar{q_a}p_uq_u}{(q_u^2 + p_u^2)^{3/2}}, 1\right] 
%    \begin{bmatrix}
%        -(1 + \cos(q_a)) \\
%        (3 + 2\cos(q_a))
%    \end{bmatrix} \\
%                     &= \frac{ (3+2\cos(q_a))(q_u^2 + p_u^2)^{3/2} 
%                     - \bar{q_a}p_uq_u(1+cos(q_a))}
%                     {m l^2 (q_u^2+p_u^2)^{3/2} (2 - \cos^2(q_a))}
%    \end{split}
%\end{align}
%
%TODO: \textbf{Determine if this is full rank everywhere except at \((q_u,p_u) = (0,0)\)}
%
%Solving for \(p_a\) is straightforward. The computation is shown below.
%\begin{multline*}
%    p_a = \left(dh_q M^{-1}(q)B\right)^{-1}  \\
%        \left(\frac{m g l \bar{q_a}p_u q_u^2}{(q_u^2+p_u^2)^{3/2}} 
%            \left(2\sin(q_u) + sin(q_u + q_a) \right)
%        +
%            \frac{1}{m l^2 (2 - \cos^2(q_a))}
%            [ \frac{\bar{q_a}p_uq_u}{(q_u^2+p_u^2)^{3/2}}, 1] 
%            \begin{bmatrix}
%              1 \\
%              -(1 + \cos(q_a))
%          \end{bmatrix} \right)
%\end{multline*}
%\begin{align*}
%    &= -\frac
%    {
%        \frac{ m g l \bar{q_a}q_u^2(2\sin(q_u) + \sin(q_u+q_a))}{(q_u^2+p_u^2)^{3/2}}
%        + \frac{\bar{q_a} p_u^2 q_u}{m l^2 (q_u^2 + p_u^2)^{3/2} (2 -\cos^2(q_a))}
%        - \frac{(1+\cos(q_a)) p_u}{m l^2 (2 - \cos^2(q_a))}
%    }
%    {
%        \frac
%        {
%            (3 + 2\cos(q_a))(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + \cos(q_a))q_up_u
%        }
%        {
%            m l^2 (q_u^2+p_u^2)^{3/2} (2 - \cos^2(q_a))
%        }
%    } \\
%    &= \frac
%    {
%        \frac
%        {
%            -m^2 g l^3 \bar{q_a}q_u^2(2\sin(q_u) + \sin(q_u+q_a))(2 - \cos^2(q_a))
%            - \bar{q_a}p_u^2 q_u 
%            + (1 + \cos(q_a))(q_u^2 + p_u^2)^{3/2} p_u
%        }
%        {
%            m l^2 (q_u^2+p_u^2)^{3/2} (2 - \cos^2(q_a))
%        }
%    }
%    {
%        \frac
%        {
%            (3 + 2\cos(q_a))(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + \cos(q_a))q_up_u
%        }
%        {
%            m l^2 (q_u^2+p_u^2)^{3/2} (2 - \cos^2(q_a))
%        }
%    }
%\end{align*}
%
%\begin{equation*}
%    \Rightarrow p_a = \frac
%        {
%            (1 + \cos(q_a))(q_u^2 + p_u^2)^{3/2} p_u
%            -m^2 g l^3 \bar{q_a}q_u^2(2\sin(q_u) + \sin(q_u+q_a))(2 - \cos^2(q_a))
%            - \bar{q_a}q_up_u^2
%        }
%        {
%            (3 + 2\cos(q_a))(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + \cos(q_a))q_up_u
%        }
%\end{equation*}
%
%With \(p_a\) in hand, it is now possible to solve for the zero dynamics under the constraint. To simplify the notation we define \(s_u := \sin(q_u)\),
%\(s_{ua} := \sin(q_u + q_a)\), 
%and \(c_a := \cos(q_a)\).
%\begin{align*}
%    \dot{q_u} &= \frac{1}{m l^2 (2 - c_a^2)} [ 1, -(1+c_a)]
%    \begin{bmatrix} p_u \\ p_a \end{bmatrix} \\
%    \dot{p_u} &= - m g l (2 s_u + s_{ua})
%\end{align*}
%Expanding out \(\dot{q_u}\) we get
%\begin{multline*}
%    \dot{q_u} = 
%    \frac{p_u}{m l^2 (2 - c_a^2)}\\
%    + \frac
%    {
%        m^2 g l^3 \bar{q_a}q_u^2(2s_u + s_{ua})(2 - c_a^2)(1 + c_a)
%        - (1+c_a)^2(q_u^2+p_u^2)^{3/2}p_u 
%        + \bar{q_a}(1+c_a)q_up_u^2
%    }
%    {
%        ml^2 (2-c_a^2) ((3 + 2c_a)(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + c_a)q_up_u)
%    }
%\end{multline*}
%and, putting everything over one denominator,
%\begin{align*}
%    \dot{q_a} &= \frac
%    {
%        (3 + 2c_a)(q_u^2+p_u^2)^{3/2}p_u 
%        - (1+c_a)^2(q_u^2+p_u^2)^{3/2}p_u 
%        +m^2 g l^3 \bar{q_a}q_u^2(2s_u + s_{ua})(2 - c_a^2)(1 + c_a)
%    }
%    {
%        ml^2 (2-c_a^2) ((3 + 2c_a)(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + c_a)q_up_u)
%    } \\
%    &= \frac
%    {
%        (2-c_a^2)\left((q_u^2+p_u^2)^{3/2}p_u 
%        +m^2 g l^3 \bar{q_a}q_u^2(2s_u + s_{ua})(1 + c_a)\right)
%    }
%    {
%        ml^2 (2-c_a^2) ((3 + 2c_a)(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + c_a)q_up_u)
%    }
%\end{align*}
%Therefore, the zero-dynamics of the acrobot under the constraint 
%\(q_a = \sin(\arctan(p_u,q_u))\) are:
%\begin{align}\label{eqn:acrobot_eom_qp}
%\begin{split}
%    \dot{q_u} &= \frac
%    {
%        (q_u^2+p_u^2)^{3/2}p_u 
%        +m^2 g l^3 \bar{q_a}q_u^2(2s_u + s_{ua})(1 + c_a)
%    }
%    {
%        ml^2 \left((3 + 2c_a)(q_u^2+p_u^2)^{3/2} - \bar{q_a}(1 + c_a)q_up_u\right)
%    } \\
%    \dot{p_u} &= - m g l (2s_u + s_{ua})
%\end{split}
%\end{align}
%
%It may be convenient to represent these equations of motion  in polar coordinates.
%Letting \(r = \sqrt{q_u^2 + p_u^2}\) be the radius and 
%\(\theta = \arctan(p_u,q_u)\) be the angle in the \((q_u,p_u)\) plane and taking
%derivatives, it is easy to show that
%\begin{align}\label{eqn:dr_and_dt_general}
%\begin{split}
%    \dot{r} &= [\cos(\theta),\sin(\theta)]
%    \begin{bmatrix} \dot{q_u}(r,\theta) \\ \dot{p_u}(r,\theta) \end{bmatrix} \\
%    \dot{\theta} &= \frac{1}{r}[-\sin(\theta), \cos(\theta)]
%    \begin{bmatrix} \dot{q_u}(r,\theta) \\ \dot{p_u}(r,\theta) \end{bmatrix}
%\end{split}
%\end{align}
%
%By expanding out (\ref{eqn:dr_and_dt_general}) then substituting \(q_u = r\cos(\theta)\)
%and \(p_u = r\sin(\theta)\)
%it is straightforward computation to find the equations of motion in cylindrical \((r,\theta)\)
%coordinates. To simplify the notation yet again, define
%\(s(g)_\theta := \sin(g(\theta))\) and \(c(g)_\theta := \cos(g(\theta))\) 
%for any function \(g: \mathbb{R} \rightarrow \mathbb{R}\). The final equations
%of motion in \((r,\theta)\) coordinates are
%\begin{align}\label{eqn:acrobot_eom_rt}
%\begin{split}
%    \dot{r} &= \frac
%    {
%        r^2 c_\theta s_\theta 
%        + m^2 g l^3 (2 s(rc)_\theta + s(rc + \bar{q_a}s)_\theta)
%        \left(
%            (1+c(\bar{q_a}s)_\theta)\bar{q_a}c_\theta 
%            - (3 + 2c(\bar{q_a}s)_\theta)s_\theta r
%        \right)
%    }
%    {
%        m l^2 \left((3+2c(\bar{q_a}s)_\theta)r 
%        - \bar{q_a}(1+ c(\bar{q_a}s)_\theta)c_\theta s_\theta\right)
%    }\\
%    \dot{\theta} &= -\frac
%    {
%        r s_\theta^2
%        + m^2 g l^3 (2 s(rc)_\theta + s(rc + \bar{q_a}s)_\theta)
%        (3 + 2c(\bar{q_a}s)_\theta)c_\theta
%    }
%    {
%        m l^2 \left((3+2c(\bar{q_a}s)_\theta)r 
%        - \bar{q_a}(1+ c(\bar{q_a}s)_\theta)c_\theta s_\theta\right)
%    }
%\end{split}
%\end{align}

>>>>

\begin{figure}
    \centering
    \includestandalone[width=0.7\textwidth]{images/acrobot_model}
    \caption{The general acrobot model, represented by two weighted rods
    differing in both length and mass.}%
    \label{fig:acrobot-model}
\end{figure}

@@

\section{Dynamics of the Acrobot}
Suppose we are given an acrobot as in Figure \ref{fig:acrobot-model} modelling a
gymnast hanging on a horizontal bar, where the ``torso" has moment of
inertia \(J_u\) and the ``leg" link has moment of inertia \(J_a\) (each with respect
to their own center of mass).
Let \(q_u \in \mathbb{S}^1\) be the shoulder angle and \(q_a \in \mathbb{S}^1\) 
be the hip angle, where only \(q_a\) is actuated. 
Collecting them together provides the configuration
\(q = (q_u,q_a) \in \mathbb{S}^1 \times \mathbb{S}^1\). 
The acrobot has inertia matrix \(D\), potential function \(P\) (with respect to
the horizontal bar), and input matrix \(B\)  given as
follows \cite{xingbo_thesis}:
\begin{align}\label{eqn:general-acrobot-inertia}
    D(q) &= \begin{bmatrix}
      m_al_u^2 + 2m_a\cos(q_a)l_u l_{c_a} + m_al_{c_a}^2 + m_ul_{c_u}^2 + J_u + J_a &
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a \\
      m_al_{c_a}^2 + m_al_ul_{c_a}\cos(q_a) + J_a &
      m_al_{c_a}^2 + J_a
    \end{bmatrix} \\
    \label{eqn:general-acrobot-potential}
    P(q) &= g\left(m_al_{c_a}(1 - \cos(q_u+q_a)) + 
        (m_al_u + m_ul_{c_u})(1-\cos(q_u))\right) \\
    B(q) &= \begin{bmatrix} 0 \\ 1 \end{bmatrix}
\end{align}

While this is the most general representation of an acrobot, the dynamics are
unwieldy to the point where it becomes difficult to prove the results in this
chapter analytically.
We will therefore simplify the dynamics by assuming the acrobot is comprised of
two massless rods of equal length \(l\), with equal point masses \(m\) at the
tips (as in Figure \ref{fig:simple-acrobot-model}).
We call this type of acrobot a \textbf{simple} acrobot. We will also ignore any
frictional forces at both the hips and shoulder. Finally, it is important to note
that a real gymnast cannot swing their hips in full circles; for this reason, we
assume that \(q_a \in [-\bar{q}_a, \bar{q}_a]\) where 
\(\bar{q}_a \in ]-\pi, \pi[\). On mechanical acrobots, this can be taken as a
control parameter.

\begin{figure}
    \centering
    \includestandalone[width=0.7\textwidth]{images/simple_acrobot_model}
    \caption{A simple acrobot has massless rods of equal length \(l\) and 
    equal masses \(m\) at the tips.}
    \label{fig:simple-acrobot-model}
\end{figure}

Given the assumptions listed, the moments of inertia \(J_u\) and \(J_a\) of the
rods vanish. Furthermore, \(l_{c_u} = l_{c_a} = l_u = l_a = l\) and
\(m_u = m_a = m\). Reducing
(\ref{eqn:general-acrobot-inertia}-\ref{eqn:general-acrobot-potential}) with this
knowledge yields the following ``simple" inertia matrix and potential function:
\begin{align}
    D(q) &= \begin{bmatrix}
        ml^2\left(3+2\cos(q_a)\right) & 
        ml^2\left(1+\cos(q_a)\right) \\
        ml^2\left(1+\cos(q_a)\right) &
        ml^2
    \end{bmatrix} \\
    P(q) &= -mgl\left(2\cos(q_u)+\cos(q_u+q_a)\right)
\end{align}
\begin{notation}
    For shorthand, we write \(c_u := \cos(q_u)\), \(c_a := \cos(q_a)\), and 
    \(c_{ua} := \cos(q_u + q_a)\). Likewise, \(s_u := \sin(q_u)\), 
    \(s_a := \sin(q_a)\), and \(s_{ua} := \sin(q_u + q_a)\).
\end{notation}

Defining \(M(q) := D(q)\) and \(V(q) := P(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)\\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
\end{equation}

\section{Previous Constraint Approaches}
% TODO: Give Xingbo's intuition, abstract away to qa = sin(theta) constraint
% based on VLP but show it doesn't work well in our framework
% TODO: Describe Xingbo's results and observe that his results were similar to a VNHC
% except they used VHC tools. We will modify his approach to use VNHCs and prove
% rigorously that we can stabilize any energy level set on the acrobot.

@@
